---
title: "Wealth and Happiness"
author: 
- "Luuk Neefjes 2898224" 
- "Ilias Abgij"
- "Denis Bykov" 
- "Finn van Echtelt" 
- "Merlijn Haster" 
- "Milan Kroes"
- "Raluca Oprina" 
- "Tutorial Group 04"
- "Fatima Amankour"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("ggplot2")
install.packages("readxl")
install.packages("knitr")
install.packages("dplyr")
install.packages("rnaturalearth")
install.packages("rnaturalearthdata")
install.packages("sf")
```

```{r package_requirement, include=FALSE}
require(tidyverse)
require(rmarkdown)
require(yaml)
library(readxl)
library(knitr)
library(tinytex)
require(dplyr)
require(rnaturalearth)
require(rnaturalearthdata)
require(sf)
```

# Part 1 - Identification of the social problem

## 1.1 Describe the Social Problem

The relation between wealth and happiness in countries is a social problem because many societies think that wealth is the key to a happy life. While richer countries tend to report higher life satisfaction, research shows that beyond a certain income level, more wealth does not equal more happiness (Kahneman & Deaton, 2010). The Easterlin paradox suggests that economic growth over time does not increase happiness at the same rate forever (Easterlin, 1974).

In this report, we examine GDP, inflation, poverty, and happiness between 2014 and 2025 in the Benelux. \# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}

poverty <- read_xlsx("data/povertybenelux.xlsx")
happiness <- read_xlsx("data/happinessbenelux.xlsx")
gdp <- read_xlsx("data/gdpbenelux.xlsx")
inflation <- read_xlsx("data/inflationbenelux.xlsx")
```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

- Do the variables contain health information or SES information?

- Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

# Steps taken for data cleaning

1.  **strip non-numeric characters**
2.  **pivot from wide to long**
3.  **convert the year column to integer**
4.  **have the GDP values also in line with the earlier observations that are already in billions**
5.  **'left-join' all four long tables on 'country' and 'year'**

```{r data_cleaning}
clean_long <- function(df, value_col) {
  df%>%
    pivot_longer(-Country,names_to = "year", values_to = value_col) %>%
    mutate(year = as.integer(year))
}
```

```{r remove_symbols}
gdp <- gdp %>%
  mutate(across(-Country, ~ round(as.numeric(gsub(",", "", .)), 3)))

inflation <- inflation %>%
  mutate(across(-Country, ~ as.numeric(gsub("%", "", .))))

poverty <- poverty %>%
  mutate(across(-Country, ~ as.numeric(gsub("%", "", .))))
```

```{r pivot_too_long}
gdp <- clean_long(gdp, "gdp")
inflation <- clean_long(inflation, "inflation")
happiness <- clean_long(happiness, "happiness")
poverty <- clean_long(poverty, "poverty")
```

```{r merge_data}
merged <- gdp %>%
  left_join(happiness, by = c("Country", "year")) %>%
  left_join(inflation, by = c("Country", "year")) %>%
  left_join(poverty, by = c("Country", "year"))
```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}
merged <- merged %>%
  mutate(period = case_when(
    year <= 2017 ~ "Early",
    year <= 2021 ~ "Mid",
    TRUE ~ "Late"
  ))
```

## 3.3 Visualize temporal variation

```{r temporal_variation}
merged %>%
  pivot_longer(cols = c(happiness, inflation, poverty),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = year, y = value, color = Country)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ variable, scales = "free_y") +
  labs(title = "Happiness, GDP(PPP), inflation, poverty Benelux",
       x = "Year",
       y = "Value",
       color = "Country")

ggplot(merged, aes(x = year, y = gdp, color = Country)) +
  geom_line() +
  geom_point() +
  scale_y_log10() +
  labs(title = "GDP (PPP) in Benelux (log scale)",
       x = "Year",
       y = "GDP(PPP)",
       color = "Country")
```

```{r gdp_variation}
```

## 3.4 Visualize spatial variation

```{r average_happiness}
avg_happiness <- merged %>%
  group_by(Country) %>%
  summarise(avg_happiness = mean(happiness, na.rm = TRUE))
```

```{r map}
world_map <- ne_countries(scale = "medium", returnclass = "sf")
```

```{r join_map_data}
benelux_map <- world_map %>%
  left_join(avg_happiness, by = c("name" = "Country"))
```

```{r spatial_variation}
ggplot(benelux_map) +
  geom_sf(aes(fill = avg_happiness)) +
  coord_sf(xlim = c(2, 7), ylim = c(49, 54), expand = FALSE) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(title = "Average happiness Benelux (2014-2025)",
       fill = "Average happiness") +
  theme_minimal()

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r boxplot}
ggplot(merged, aes(x = period, y = happiness, fill = period)) + geom_boxplot() +
  facet_wrap(~ Country) + 
  labs(title = "Happiness by time period, split by country",
       x = "Time period",
       y = "Happiness score")
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}
merged %>%
  pivot_longer(cols = c(happiness, inflation),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = year, y = value, color = Country)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "red") +
  facet_wrap(~ variable, scales = "free_y") +
  labs(title = "Happines & inflation with COVID",
       x = "Year",
       y = "Value",
       color = "Country")
```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

-Easterlin, R.A. (1974). Does economic growth improve the human lot? In P.A. David & M.W. Reder (EDS.), nations and households in economic growth (pp. 89-125). Academic press.

-Kahneman, D. & Deaton, A. (2010). High income improves evaluation of life but not emotional well-being. Proceedings of the national academy of sciences,107(38), 16489-16493. <https://doi.org/10.1073/pnas.1011492107>

-Eurostat. (n.d.). At-risk-of-poverty rate by poverty threshold, age and sex (tespm010). European Commission. Retrieved June 18, 2026, from Eurostat Data Browser

-StatBase. (n.d.). Happiness index: Netherlands – yearly data, chart and table. Retrieved June 18, 2026, from https://statbase.org/data/nld-happiness-index/

-StatBase. (n.d.). Happiness index: Belgium – yearly data, chart and table. Retrieved June 18, 2026, from https://statbase.org/data/bel-happiness-index/ 

-StatBase. (n.d.). Happiness index: Luxembourg – yearly data, chart and table. Retrieved June 18, 2026, from https://statbase.org/data/lux-happiness-index/ 

-Eurostat. (n.d.). Price level indices (tec00120). European Commission. Retrieved June 18, 2026, from https://ec.europa.eu/eurostat/databrowser/view/tec00120/default/table?lang=en

-Eurostat. (n.d.). HICP – inflation rate (tec00118). European Commission. Retrieved June 18, 2026, from https://ec.europa.eu/eurostat/databrowser/view/tec00118/default/table?lang=en
