---
title: "Wealth and Happiness"
author: 
- "Luuk Neefjes 2898224" 
- "Ilias Abgij 2697642"
- "Denis Bykov" 
- "Finn van Echtelt 2871366"
- "Merlijn Haster 2903322" 
- "Milan Kroes 2890388"
- "Raluca Oprina 2878997" 
- "Tutorial Group 04"
- "Fatima Amankour"
date: "`r Sys.Date()`"
output: pdf_document
toc: true
---


```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("ggplot2")
install.packages("readxl")
install.packages("knitr")
install.packages("dplyr")
install.packages("rnaturalearth")
install.packages("rnaturalearthdata")
install.packages("sf")
```

```{r package_requirement, include=FALSE}
require(tidyverse)
require(rmarkdown)
require(yaml)
library(readxl)
library(knitr)
library(tinytex)
require(dplyr)
require(rnaturalearth)
require(rnaturalearthdata)
require(sf)
```

# 1 - Identification of the social problem

## 1.1 Describe the Social Problem

The relation between wealth and happiness in countries is a social problem because many societies think that wealth is the key to a happy life. While richer countries tend to report higher life satisfaction, research shows that beyond a certain income level, more wealth does not equal more happiness (Kahneman & Deaton, 2010). The Easterlin paradox suggests that economic growth over time does not increase happiness at the same rate forever (Easterlin, 1974). More recently, researchers found that happiness does continue to rise with income for most people, but not for those who are already experiencing emotional distress. (Killingsworth et al. 2023)The relation between wealth and happiness is dependent on the context. Most studies focus on individual level income more than macro level indicators such as poverty and inflation.

In this report, we address that gap by researching the GDP, inflation, poverty, and happiness between 2014 and 2025 in the Benelux.

# 2 - Data Sourcing

## 2.1 Load in the data

```{r loading_data}

poverty <- read_xlsx("data/povertybenelux.xlsx")
happiness <- read_xlsx("data/happinessbenelux.xlsx")
gdp <- read_xlsx("data/gdpbenelux.xlsx")
inflation <- read_xlsx("data/inflationbenelux.xlsx")
```

## 2.2 Provide a short summary of the dataset(s)

This project uses four datasets, covering data from Belgium, Luxembourg, and the Netherlands from 2014-2025:

GDP contains GDP per capita expressed in purchasing power parity (index, EU average = 100), sourced from Eurostat. It reflects how wealthy a country is relative to the EU average, adjusting for price level differences. A limitation is that the relative measurement is sensitive to shifts in other EU countries. 

Happiness contains annual happiness scores (scale 0–10) per country, sourced from StatBase, which aggregates World Happiness Report data. The score reflects residents' subjective life satisfaction. A limitation is that this aggregation relies on self-reported life satisfaction and may not be consistent in research across years. 

Inflation contains the annual Harmonised Index of Consumer Prices (HICP) inflation rate (%), sourced from Eurostat. It captures the change in the general price level that households face.

Poverty contains the at-risk-of-poverty rate (% of population below 60% of median income), sourced from Eurostat. It serves as a measure of how many income groups are at risk of living in poverty in a country. A limitation is that it captures inequality rather than absolute deprivation, so a country can improve on this metric by narowwing the gap between rich and poor rather than lifting the poorest. 

All four datasets have the same limitation, they are annual national averages. Variation by region, age, or income group is not visible in these datasets.

All four datasets were structured in wide format (countries as rows, years as columns) and were cleaned and merged into a single long-format dataset for analysis. 

# 3 - Quantifying

## 3.1 Data cleaning

The raw datasets needed several cleaning steps before they could be merged and analysed. 

First, commas were stripped from GDP values and percentage signs were stripped from inflation and poverty figures, converting all columns to numeric. Without this step, R would treat these as character strings and all calculations would fail. And we don't want that.

Second, all four datasets were pivoted from wide format (one column per year) to long format (one row per country-year), which is required for ggplot2 and for the join.

Third, the year column was converted from character to integer so it can be used on a continuous axis. 

Finally, all four long tables were merged using **left_join()** on Country and year, keeping all GDP rows as the base and linking them with happiness, inflation, and poverty values. A disadvantage of the left join is that if a country-year exists in GDP but not in another dataset, the row is retained with an NA — this preserves data completeness but means some analyses will implicitly drop those rows via **na.rm = TRUE**. No rows were manually removed; all missing values are handled at the analysis stage. 


```{r remove_symbols}
gdp <- gdp %>%
  mutate(across(-Country, ~ round(as.numeric(gsub(",", "", .)), 3)))

inflation <- inflation %>%
  mutate(across(-Country, ~ as.numeric(gsub("%", "", .))))

poverty <- poverty %>%
  mutate(across(-Country, ~ as.numeric(gsub("%", "", .))))
```

```{r data_cleaning}
clean_long <- function(df, value_col) {
  df%>%
    pivot_longer(-Country,names_to = "year", values_to = value_col) %>%
    mutate(year = as.integer(year))
}
```

```{r pivot_too_long}
gdp <- clean_long(gdp, "gdp")
inflation <- clean_long(inflation, "inflation")
happiness <- clean_long(happiness, "happiness")
poverty <- clean_long(poverty, "poverty")
```

```{r merge_data}
merged <- gdp %>%
  left_join(happiness, by = c("Country", "year")) %>%
  left_join(inflation, by = c("Country", "year")) %>%
  left_join(poverty, by = c("Country", "year"))
```

## 3.2 Describe the type of variables included

The variables in this project are macro-level socioeconomic and well-being indicators collected at the country-year level. GDP (PPP) and inflation are sourced from administrative and national accounts through Eurostat. The poverty rate is based on household survey data (EU-SILC) aggregated to the national level. The happiness score is based on survey responses in which individuals rate their life satisfaction, that is subjective data.

## 3.3 Generate necessary variables

We construct two new variables to enable sub-population comparison.

**period** bins the continuous year variable into three categorical time periods: Early (2014–2017), Mid (2018–2021), and Late (2022–2025). This allows us to compare happiness distributions across distinct economic phases rather than tracking individual years, which is useful for the boxplot in Section 3.5. 

**happiness_change** captures year-on-year change in happiness per country, calculated as happiness - lag(happiness) within each country group. This variable adds analytical value by showing whether happiness is actively rising or falling in a given year, rather than just its absolute level — which is relevant when examining the impact of the 2022 inflation shock. 

```{r gen_var1}
merged <- merged %>%
  mutate(period = case_when(
    year <= 2017 ~ "Early",
    year <= 2021 ~ "Mid",
    TRUE ~ "Late"
  ))
```

```{r gen_var2}
merged <- merged %>%
  group_by(Country) %>%
  arrange(year) %>%
  mutate(happiness_change = happiness - lag(happiness)) %>%
  ungroup() 
```

## 3.4 Visualize temporal variation

```{r temporal_variation}
merged %>%
  pivot_longer(cols = c(happiness, inflation, poverty),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = year, y = value, color = Country)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ variable, scales = "free_y") +
  labs(title = "Happiness, GDP(PPP), inflation, poverty Benelux",
       x = "Year",
       y = "Indicator value",
       color = "Country",
       caption = "Happiness 0-10 scale. Inflation: HICP annual rate (%). Poverty: at-risk-of-poverty rate (%).")

ggplot(merged, aes(x = year, y = gdp, color = Country)) +
  geom_line() +
  geom_point() +
  scale_y_log10() +
  labs(title = "GDP (PPP) in Benelux (log scale)",
       subtitle = "GDP per capita as index relative to EU average (EU = 100)",
       x = "Year",
       y = "GDP(PPP)",
       color = "Country")
```

Happiness scores remained pretty stable across all three countries from 2014 to 2025. Inflation was low until spiking sharply in 2022 due to the post-COVID energy crisis, then fell again. Despite these shocks, happiness did not drop quickly, supporting the Easterlin paradox.

## 3.5 Visualize spatial variation

```{r average_happiness}
avg_happiness <- merged %>%
  group_by(Country) %>%
  summarise(avg_happiness = mean(happiness, na.rm = TRUE))
```

```{r map}
world_map <- ne_countries(scale = "medium", returnclass = "sf")
```

```{r join_map_data}
benelux_map <- world_map %>%
  left_join(avg_happiness, by = c("name" = "Country"))
```

```{r spatial_variation}
ggplot(benelux_map) +
  geom_sf(aes(fill = avg_happiness)) +
  coord_sf(xlim = c(2, 7), ylim = c(49, 54), expand = FALSE) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(title = "Average happiness score by Benelux country (2014–2025)",
     subtitle = "Happiness measured on a 0–10 scale ",
     fill = "Avg. happiness (0–10 scale)") +
  theme_minimal()

```

The map shows the Netherlands had the highest average happiness over 2014–2025, followed by Luxembourg, then Belgium. If we look at Luxembourg it has by far the highest GDP and yet does not rank first in happiness, supporting the idea that wealth beyond a certain point does not equal greater life satisfaction.

## 3.6 Visualize sub-population variation

```{r boxplot}
ggplot(merged, aes(x = period, y = happiness, fill = period)) + geom_boxplot() +
  facet_wrap(~ Country) + 
  labs(title = "Happiness by time period, split by country",
       x = "Time period (Early: 2014-2017, Mid: 2018-2021, Late: 2022-2025",
       y = "Happiness score (0-10 scale)",
       fill = "period")
```

We can see that happiness was stable across time periods in Belgium, while Luxembourg and the Netherlands showed a slight rise into the Late period. The Netherlands scored highest in every period. This suggests that factors beyond GDP — such as inequality and social policy — are also important factors for overall well-being.

## 3.7 Event analysis

```{r analysis}
merged %>%
  pivot_longer(cols = c(happiness, inflation),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = year, y = value, color = Country)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "red") +
  annotate("text", x = 2020.2, y = Inf, label = "COVID-19\npandemic", 
           hjust = 0, vjust = 1.5, size = 3, color = "red") +
  facet_wrap(~ variable, scales = "free_y", ncol = 1) +
  labs(title = "Happiness & inflation around the COVID-19 pandemic (Benelux)",
     x = "Year",
     y = "Indicator value",
     color = "Country",
     caption = "Happiness: 0–10 scale. Inflation: HICP annual rate (%). Red dashed line = COVID-19 pandemic (2020).")
```
Be20, inflation was low and happiness was steady across all three countries. After the COVID shock, inflation spiked dramatically in 2022 while happiness plateaued or declined slightly. This co-movement suggests inflation may dampen well-being through reduced purchasing power, though causality cannot be established from this association alone.

# Part 4 - Discussion

The relationship between wealth and happiness was studied in Benelux from 2014 to 2025. GDP, inflation, poverty and happiness scores were used as key variables for this research. Numerous studies and analyses provide the rationale for this thesis. Easterlin (1974) demonstrated how richer people are always happier than poorer ones in a country; but it also showed that rising national income does not lead to rising happiness over time. This statement has been widely discussed in the past. The contentment levels of most individuals still increase with their income, Killingsworth et al. (2023) recently showed. However, this link is weaker for individuals that are already suffering from depression, suggesting that one’s wealth doesn’t necessarily equal happiness. This paper utilises the COVID-19 pandemic (2020–2021) and its economic impacts particularly the increase in inflation in 2022 as a natural experiment to study the relationship between the real economy and the financial markets.

In general, the fluctuation of wealth, due to the ups and downs happening in the economy, never makes one immune to it. Happiness reduces significantly when it noted that the inflation spikes in 2022 and the 2020 Covid shock, which was when prices and purchasing power became more than reasonable expectations. This is also the case for rich countries, note. The literature also mentions a match of this sort. Killingsworth et al. (2023) the observations reflect that wealth is not always a guaranteed factor of happiness. In addition, the reminder of Easterlin’s (1974) paradox is that it is not enough for the economy of the society to grow for happiness to grow. After we pass the threshold, wealth stability and distribution rather than its magnitude will determine happiness.

# Part 5 - Reproducibility

## 5.1 Github repository link

<https://github.com/luukneefjes-prog/happiness>

## 5.2 Reference list

-Easterlin, R.A. (1974). Does economic growth improve the human lot? In P.A. David & M.W. Reder (EDS.), nations and households in economic growth (pp. 89-125). Academic press.

-Kahneman, D. & Deaton, A. (2010). High income improves evaluation of life but not emotional well-being. Proceedings of the national academy of sciences,107(38), 16489-16493. <https://doi.org/10.1073/pnas.1011492107>

-Eurostat. (n.d.). At-risk-of-poverty rate by poverty threshold, age and sex (tespm010). European Commission. Retrieved June 18, 2026, from Eurostat Data Browser

-StatBase. (n.d.). Happiness index: Netherlands – yearly data, chart and table. Retrieved June 18, 2026, from <https://statbase.org/data/nld-happiness-index/>

-StatBase. (n.d.). Happiness index: Belgium – yearly data, chart and table. Retrieved June 18, 2026, from <https://statbase.org/data/bel-happiness-index/>

-StatBase. (n.d.). Happiness index: Luxembourg – yearly data, chart and table. Retrieved June 18, 2026, from <https://statbase.org/data/lux-happiness-index/>

-Eurostat. (n.d.). Price level indices (tec00120). European Commission. Retrieved June 18, 2026, from <https://ec.europa.eu/eurostat/databrowser/view/tec00120/default/table?lang=en>

-Eurostat. (n.d.). HICP – inflation rate (tec00118). European Commission. Retrieved June 18, 2026, from <https://ec.europa.eu/eurostat/databrowser/view/tec00118/default/table?lang=en>

-Killingsworth, M. A., Kahneman, D., & Mellers, B. (2023). Income and emotional well-being: A conflict resolved. Proceedings of the National Academy of Sciences, 120(10), e2208661120. <https://doi.org/10.1073/pnas.2208661120>
